---
title: "67-metabolite_spearman"
title-block-banner: true
author:
  - name: Shujie Tian
date: 2025-04-28
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format: 
  html:
    embed-resources: true
    smooth-scroll: true
    page-layout: full
reference-location: section
citation-location: document
params:
  name: "67-metabolite_spearman"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is ...

```{r setup}
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
params <- list(name = "67-metabolite_spearman")
here::i_am(paste0(params$name, ".qmd"), uuid = "bb3547d9-67c2-4cae-a575-6c27012330ea")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r packages}
#| message: false
#| warning: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(data.table)
  library(Hmisc)
  library(pheatmap)
  library(ggsignif)
  devtools::load_all()
})
```

## Tasks

The first task is ...



```{r }
unique_metasbolit <- read.csv(path_source("66-metabolite","data_normalized221_metabolites32_name_t_group_final_t.csv"))

a37 <- openxlsx::read.xlsx(here(path_data, "37.xlsx"))
unique_bacteria <- read.csv(here(path_raw, "d61-spearman/unique.csv"))

# Process the data
a37 <- a37 %>%
  mutate(
    X = paste0("A", sample_id),
    
    # Remove all "-" from the third column
    col3 = str_remove_all(wing_no, "-")
  )

# Iterate through col1 column of df1
for (i in 1:nrow(a37)) {
  old_col_name <- a37$col3[i]  # Get value from col1 in df1
  new_col_name <- a37$X[i]     # Get value from X column as new column name
  if (old_col_name %in% names(unique_bacteria)) {  # Check if it's a column name in df2
    names(unique_bacteria)[names(unique_bacteria) == old_col_name] <- new_col_name  # Modify column name
  }
}

# Extract common column names between the two tables (excluding first column of df1)
common_columns <- dplyr::intersect(names(unique_bacteria)[-1], names(unique_metasbolit))
print(common_columns)
#A352 

# Keep first column of df1 and common columns
unique_bacteria32 <- unique_bacteria[c("sample_no", common_columns)]

unique_metasbolit <- unique_metasbolit[c("X", common_columns)]

# Transpose unique_metasbolit
meta437 <- unique_metasbolit$X
unique_metasbolit_t <- as.data.frame(t(unique_metasbolit))
unique_metasbolit_t <- unique_metasbolit_t[-1,]
names(unique_metasbolit_t) <- meta437

# Transpose unique_bacteria32
bac21 <- unique_bacteria32$sample_no
unique_bacteria32_t <- as.data.frame(t(unique_bacteria32))
unique_bacteria32_t <- unique_bacteria32_t[-1,]
names(unique_bacteria32_t) <- bac21

# Match row order between the two transposed data frames
unique_metasbolit_t <- unique_metasbolit_t[match(row.names(unique_bacteria32_t), row.names(unique_metasbolit_t)),]
```



#metasbolit_fcr all
```{r }
# Load and prepare character data
character_new <- read.csv(here(path_raw, "d61-spearman/character.csv"), header = T, row.names = 1)
character_t <- t(character_new)
character_t <- as.data.frame(character_t[c("FCR","ADG"),])

# Rename columns based on a37 mapping
for (i in 1:nrow(a37)) {
  old_col_name <- a37$col3[i]   
  new_col_name <- a37$X[i]  
  if (old_col_name %in% names(character_t)) {  
    names(character_t)[names(character_t) == old_col_name] <- new_col_name  # Modify column names
  }
}

# Find common columns between datasets
common_columns <- dplyr::intersect(names(character_t), row.names(unique_metasbolit_t))
print(common_columns)

# Filter and transpose character data to match metabolite data
character_t31 <- character_t[,common_columns]
character_t31 <- as.data.frame(t(character_t31))
character_t31 <- character_t31[row.names(unique_metasbolit_t),]

# 1. Merge data frames
metasbolits_bacteria <- cbind(unique_metasbolit_t, character_t31)

# Calculate correlations and significance
spearman_results <- rcorr(as.matrix(metasbolits_bacteria), type = "spearman")
spearman_corr <- spearman_results$r
spearman_pval <- spearman_results$P

# Filter correlation matrices to exclude character variables
spearman_corr <- spearman_corr[-(1:221), -(222:223)]
spearman_pval <- spearman_pval[-(1:221), -(222:223)]
spearman_corr <- t(spearman_corr)
spearman_pval <- t(spearman_pval)

# Save correlation results
write.table(spearman_corr, path_target("spcor_metasbolits_character.txt"), quote = F, row.names = T)
write.table(spearman_pval, path_target("spp_metasbolits_character.txt"), quote = F, row.names = T)
write.csv(spearman_corr, path_target("spcor_metasbolits_character.csv"), row.names = T)
write.csv(spearman_pval, path_target("spp_metasbolits_character.csv"), row.names = T)

# 2. Filter significant correlations (p < 0.05 & |r| > 0.3)
sig_corr <- spearman_corr
sig_corr[spearman_pval >= 0.05 | abs(spearman_corr) < 0.3] <- NA

# 3. Create significance marker matrix
sig_labels <- matrix("", nrow = nrow(spearman_pval), ncol = ncol(spearman_pval))
sig_labels[spearman_pval < 0.05 & abs(spearman_corr) >= 0.3] <- "*"
sig_labels[spearman_pval < 0.01 & abs(spearman_corr) >= 0.3] <- "**"
sig_labels[spearman_pval < 0.001 & abs(spearman_corr) >= 0.3] <- "***"

# 4. Filter rows and columns with at least one significant correlation
keep_rows <- rowSums(!is.na(sig_corr)) > 0
keep_cols <- colSums(!is.na(sig_corr)) > 0
sig_corr_filtered <- sig_corr[keep_rows, keep_cols]
sig_labels_filtered <- sig_labels[keep_rows, keep_cols]
sig_labels_filtered <- as.matrix(sig_labels_filtered)
filtered <- row.names(sig_corr_filtered)
sig_corr_filtered_final <- spearman_corr[filtered,]
sig_corr_filtered_final <- as.matrix(sig_corr_filtered_final)

# Prepare filtered metabolite data
unique_metasbolit39 <- unique_metasbolit[unique_metasbolit$X %in% row.names(sig_corr_filtered),]
row.names(unique_metasbolit39) <- unique_metasbolit39$X
unique_metasbolit39$X <- NULL
write.csv(unique_metasbolit39, path_target("unique_metasbolit39.csv"))

# Transpose and add group information
unique_metasbolit39_t <- as.data.frame(t(unique_metasbolit39))
unique_metasbolit39_t <- rownames_to_column(unique_metasbolit39_t, "Sample")

group32 <- read.csv(here(path_raw, "d66-metabolite/group32.csv"))
unique_metasbolit39_t_group <- merge(unique_metasbolit39_t, group32, by = "Sample", all.x = TRUE)

# Reorder columns: move last column (Group) to second position
unique_metasbolit39_t_group <- unique_metasbolit39_t_group %>%
  dplyr::select(1, ncol(unique_metasbolit39_t_group), 2:(ncol(unique_metasbolit39_t_group) - 1))
```



# Get metabolite classification information
```{r }
metabolit221_full <-  read.csv(path_source("66-metabolite", "metabolit221_full.csv"))
unique_metasbolit39_full <- metabolit221_full %>% dplyr::filter(name %in% row.names(unique_metasbolit39))
```

# Remove unimportant metabolites (FCR ADG)
```{r }
unique_metasbolit39_drug_no <- unique_metasbolit39_t[, !(names(unique_metasbolit39_t) %in% c("Piribedil","Prodigiosin","ZOLAZEPAM", "Clarithromycin", "Tezacitabine", "Difluorodeoxycytidine", "Metanephrine", "Chromoionophore I","2-(4-Hydroxy-1,3-thiazol-3-ium-3-yl)ethyl [2-(octadecylcarbamoyloxymethyl)oxolan-2-yl]methyl phosphate","1-[2,3-Dimethyl-2-(2-methylbut-3-en-2-yl)furan-3-yl]-3,5-dihydroxy-4-methoxyhexan-2-one"))]


# Data preparation ------------------------------------------------------------
row <- unique_metasbolit39_drug_no$Sample
row.names(unique_metasbolit39_drug_no) <- row
unique_metasbolit39_drug_no <- subset(unique_metasbolit39_drug_no, select = -Sample)
metab_data <- unique_metasbolit39_drug_no
microb_data <- character_t31
metab_data <- metab_data[row.names(microb_data), ]
unique_metasbolit28_full <- unique_metasbolit39_full[unique_metasbolit39_full$name %in% names(unique_metasbolit39_drug_no),]


# Metabolite classification processing
class <- unique_metasbolit28_full$super_class.HMDB.
table(class)
metab_class <- ifelse(
  class %in% c(
    "Benzenoids",
    "Lipids and lipid-like molecules",
    "Nucleosides, nucleotides, and analogues",
    "Organic acids and derivatives",
    "Organic oxygen compounds",
    "Organoheterocyclic compounds",
    "Phenylpropanoids and polyketides"
  ),
  class,
  "Others"
)

# Calculate correlations ----------------------------------------------------------
cor_res <- rcorr(as.matrix(metab_data), as.matrix(microb_data), type = "spearman")
cor_matrix <- cor_res$r[1:37, 38:39]  # Metabolites × FCR ADG
p_matrix <- cor_res$P[1:37, 38:39]

# Filter criteria: |r| > 0.3 and p < 0.05
significant_mask <- (abs(cor_matrix) > 0.3) & (p_matrix < 0.05)

# Keep rows (metabolites) with at least one significant correlation
keep_rows <- rowSums(significant_mask, na.rm = TRUE) > 0

# Keep columns (bacterial species) with at least one significant correlation
keep_cols <- colSums(significant_mask, na.rm = TRUE) > 0

# Apply filtering
cor_matrix_filtered <- cor_matrix[keep_rows, keep_cols, drop = FALSE]
p_matrix_filtered <- p_matrix[keep_rows, keep_cols, drop = FALSE]
metab_class_filtered <- metab_class[keep_rows]

# Sort by classification
class_order <- order(metab_class_filtered)
cor_matrix_sorted <- cor_matrix_filtered[class_order, , drop = FALSE]
p_matrix_sorted <- p_matrix_filtered[class_order, , drop = FALSE]

# Create significance marker matrix
sig_matrix <- matrix("", nrow = nrow(p_matrix_sorted), ncol = ncol(p_matrix_sorted))
sig_matrix[p_matrix_sorted < 0.001] <- "***"
sig_matrix[p_matrix_sorted >= 0.001 & p_matrix_sorted < 0.01] <- "**"
sig_matrix[p_matrix_sorted >= 0.01 & p_matrix_sorted < 0.05] <- "*"
sig_matrix_sorted <- sig_matrix[class_order, ]

# Annotation and color settings -----------------------------------------------------
annotation_row <- data.frame(Class = metab_class_filtered[class_order])
row.names(annotation_row) <- row.names(cor_matrix_sorted)  

class_colors <- RColorBrewer::brewer.pal(length(unique(metab_class_filtered)), "Set2")
names(class_colors) <- unique(metab_class_filtered)
anno_colors <- list(Class = class_colors)

# Draw heatmap (metabolites in rows, bacterial species in columns) ------------------------------------
p <- pheatmap::pheatmap(
  cor_matrix_sorted,  
  display_numbers = sig_matrix_sorted,
  number_color = "black",
  fontsize_number = 16,
  cluster_rows = FALSE,  
  cluster_cols = TRUE,   
  annotation_row = annotation_row,
  annotation_colors = anno_colors,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  breaks = seq(-1, 1, length.out = 101),
  main = "Significant Spearman Correlations(Metabolites vs FCR and ADG)\n(|r| ≥ 0.3,*: p<0.05, **: p<0.01, ***: p<0.001)",
  fontsize_col = 16,     
  fontsize_row = 16,    
  fontsize = 16,
  angle_col = 0,       
  border_color = "gray60",
  cellwidth = 200,       
  cellheight = 25,       
  treeheight_col = 0,   
  show_rownames = TRUE,
  show_colnames = TRUE
)

# Save figures ----------------------------------------------------------
ggsave(
  path_target("metabolite37_vs_tFCRADG_correlation.png"),
  plot = p,
  width = 24.5,  
  height = 14, 
  dpi = 600,
  bg = "white"
)
ggsave(
  path_target("metabolite37_vs_tFCRADG_correlation.pdf"),
  plot = p,
  width = 24.5,  
  height = 14,
  bg = "white"
)
```



#Perform Spearman correlation between 37 metabolites significantly associated with FCR/ADG and differential bacteria (threshold r > 0.5)
```{r }
microb_data <- unique_bacteria32_t

# Calculate correlations ----------------------------------------------------------
cor_res <- rcorr(as.matrix(metab_data), as.matrix(microb_data), type = "spearman")
cor_matrix <- cor_res$r[1:37, 38:58]  # Metabolites × Bacteria
p_matrix <- cor_res$P[1:37, 38:58]

# Filter criteria: |r| > 0.5 and p < 0.05
significant_mask <- (abs(cor_matrix) > 0.5) & (p_matrix < 0.05)

# Keep rows (metabolites) with at least one significant correlation
keep_rows <- rowSums(significant_mask, na.rm = TRUE) > 0

# Keep columns (bacteria) with at least one significant correlation
keep_cols <- colSums(significant_mask, na.rm = TRUE) > 0

# Apply filtering
cor_matrix_filtered <- cor_matrix[keep_rows, keep_cols, drop = FALSE]
p_matrix_filtered <- p_matrix[keep_rows, keep_cols, drop = FALSE]
metab_class_filtered <- metab_class[keep_rows]

# Sort by classification
class_order <- order(metab_class_filtered)
cor_matrix_sorted <- cor_matrix_filtered[class_order, , drop = FALSE]

# Create significance marker matrix
sig_matrix <- matrix("", nrow = nrow(p_matrix_filtered), ncol = ncol(p_matrix_filtered))
sig_matrix[p_matrix_filtered < 0.001] <- "***"
sig_matrix[p_matrix_filtered >= 0.001 & p_matrix_filtered < 0.01] <- "**"
sig_matrix[p_matrix_filtered >= 0.01 & p_matrix_filtered < 0.05] <- "*"
sig_matrix_sorted <- sig_matrix[class_order, ]

# Transpose matrices
cor_matrix_t <- as.data.frame(t(cor_matrix_sorted))
sig_matrix_t <- as.data.frame(t(sig_matrix_sorted))

# Annotation and color settings -----------------------------------------------------
annotation_col <- data.frame(Class = metab_class_filtered[class_order])
row.names(annotation_col) <- names(cor_matrix_t)

class_colors <- RColorBrewer::brewer.pal(length(unique(metab_class_filtered)), "Set2")
names(class_colors) <- unique(metab_class_filtered)
anno_colors <- list(Class = class_colors)

# Draw heatmap ----------------------------------------------------------
p <- pheatmap::pheatmap(
  cor_matrix_t,
  display_numbers = sig_matrix_t,
  number_color = "black",
  fontsize_number = 8,
  cluster_rows = TRUE,   
  cluster_cols = FALSE,  
  annotation_col = annotation_col,
  annotation_colors = anno_colors,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  breaks = seq(-1, 1, length.out = 101),
  cutree_rows = 1,       
  main = "",
  fontsize_col = 10,     
  fontsize_row = 10,     
  fontsize = 12,         
  angle_col = 315,       
  border_color = "gray60", 
  cellwidth = 15,        
  cellheight = 15,      
  treeheight_row = 0,   
  treeheight_col = 0,    
  show_rownames = TRUE,  
  show_colnames = TRUE   
)

# Save figures ----------------------------------------------------------
ggsave(
  path_target("metabolite37_bacteria0.5_correlation_filtered.png"),
  plot = p,
  width = 13,
  height = 9,
  dpi = 600,
  bg = "white"
)
ggsave(
  path_target("metabolite37_bacteria0.5_correlation_filtered.pdf"),
  plot = p,
  width = 13,
  height = 9,
  bg = "white"
)
```


# Perform Spearman correlation (threshold r > 0.5) between 37 metabolites significantly associated with FCR/ADG and differential VC

```{r }
unique_vc <- read.csv(here(path_data, "unique_vc.csv"), row.names = 1)
microb_data <- t(unique_vc)

# Calculate correlations ----------------------------------------------------------
cor_res <- rcorr(as.matrix(metab_data), as.matrix(microb_data), type = "spearman")
cor_matrix <- cor_res$r[1:37, 38:116]  # Metabolites × Differential VC
p_matrix <- cor_res$P[1:37, 38:116]

# Filter criteria: |r| > 0.5 and p < 0.05
significant_mask <- (abs(cor_matrix) > 0.5) & (p_matrix < 0.05)

# Keep rows with at least one significant correlation
keep_rows <- rowSums(significant_mask, na.rm = TRUE) > 0

# Keep columns with at least one significant correlation
keep_cols <- colSums(significant_mask, na.rm = TRUE) > 0

# Apply filtering
cor_matrix_filtered <- cor_matrix[keep_rows, keep_cols, drop = FALSE]
p_matrix_filtered <- p_matrix[keep_rows, keep_cols, drop = FALSE]
metab_class_filtered <- metab_class[keep_rows]

# Sort by classification
class_order <- order(metab_class_filtered)
cor_matrix_sorted <- cor_matrix_filtered[class_order, , drop = FALSE]

# Create significance marker matrix
sig_matrix <- matrix("", nrow = nrow(p_matrix_filtered), ncol = ncol(p_matrix_filtered))
sig_matrix[p_matrix_filtered < 0.001] <- "***"
sig_matrix[p_matrix_filtered >= 0.001 & p_matrix_filtered < 0.01] <- "**"
sig_matrix[p_matrix_filtered >= 0.01 & p_matrix_filtered < 0.05] <- "*"
sig_matrix_sorted <- sig_matrix[class_order, ]

# Transpose matrices
cor_matrix_t <- as.data.frame(t(cor_matrix_sorted))
sig_matrix_t <- as.data.frame(t(sig_matrix_sorted))

# Annotation and color settings -----------------------------------------------------
annotation_col <- data.frame(Class = metab_class_filtered[class_order])
rownames(annotation_col) <- names(cor_matrix_t)

class_colors <- RColorBrewer::brewer.pal(length(unique(metab_class_filtered)), "Set2")
names(class_colors) <- unique(metab_class_filtered)
anno_colors <- list(Class = class_colors)

# Draw heatmap ----------------------------------------------------------
p <- pheatmap::pheatmap(
  cor_matrix_t,
  display_numbers = sig_matrix_t,
  number_color = "black",
  fontsize_number = 8,
  cluster_rows = TRUE,   
  cluster_cols = FALSE,  
  annotation_col = annotation_col,
  annotation_colors = anno_colors,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  breaks = seq(-1, 1, length.out = 101),
  cutree_rows = 1,      
  main = "",
  fontsize_col = 10,    
  fontsize_row = 10,    
  fontsize = 12,         
  angle_col = 315,       
  border_color = "gray60", 
  cellwidth = 15,       
  cellheight = 15,      
  treeheight_row = 0,  
  treeheight_col = 0     
)

# Save figures ----------------------------------------------------------
ggsave(
  path_target("metabolite37_vc0.5_correlation_filtered.png"),
  plot = p,
  width = 14,
  height = 15,
  dpi = 600,
  bg = "white"
)
ggsave(
  path_target("metabolite37_vc0.5_correlation_filtered.pdf"),
  plot = p,
  width = 14,
  height = 15,
  bg = "white"
)
```

#ALDH5A1 / Succinic acid semialdehyde
```{r }
# Load rlog transformed data
res6_rlog <- read.csv(path_source("60-import","res6_rlog.csv"), row.names = 1)

# Prepare ALDH5A1 expression data
d <- as.data.frame(t(res6_rlog["gene-ALDH5A1",]))
d$Group <- c("ZF","ZF","ZF","CF","CF","CF")

# Define unified plotting function
plot_unified <- function(data, y_var, y_label) {
  # Ensure Group is a factor with correct levels
  data$Group <- factor(data$Group)
  
  # Calculate y-axis maximum for significance marker positioning
  y_max <- max(data[[y_var]], na.rm = TRUE)
  
  ggplot(data, aes(x = Group, y = .data[[y_var]], fill = Group)) +
    geom_boxplot(alpha = 0.5, 
                 outlier.shape = 21, outlier.size = 2) +
    # Significance testing
    geom_signif(
      comparisons = list(levels(data$Group)),  
      map_signif_level = function(p) "**",    
      y_position = y_max * 1.02,
      tip_length = 0,
      textsize = 18,
      size = 1.3
    ) +
    scale_y_continuous(limits = c(NA, y_max * 1.04)) +
    theme_bw() +
    labs(x = "", y = y_label) +
    theme(
      axis.title = element_text(size = 21),
      axis.text = element_text(size = 18),
      legend.position = "none"
    )
}

# Plot and save ALDH5A1 expression
p_ALDH5A1 <- plot_unified(d, "gene-ALDH5A1", "Abundance of ALDH5A1 (rlog)")
ggsave(p_ALDH5A1, filename = path_target("ALDH5A1.png"), width = 6, height = 6)
ggsave(p_ALDH5A1, filename = path_target("ALDH5A1.pdf"), width = 6, height = 6)

# Prepare Succinic acid semialdehyde data
unique_metasbolit_t$Sample <- row.names(unique_metasbolit_t)
unique_metasbolit_t_group_fi <- merge(unique_metasbolit_t, unique_metasbolit39_t_group, by = "Sample")
Succinic_acid_semialdehyde <- unique_metasbolit_t_group_fi[, c("Succinic acid semialdehyde", "Group")]
Succinic_acid_semialdehyde$`Succinic acid semialdehyde` <- as.numeric(Succinic_acid_semialdehyde$`Succinic acid semialdehyde`)

# Update plotting function for Succinic acid semialdehyde
plot_unified <- function(data, y_var, y_label) {
  data$Group <- factor(data$Group)
  
  # Calculate y-axis maximum for significance marker positioning
  y_max <- max(data[[y_var]], na.rm = TRUE)
  
  ggplot(data, aes(x = Group, y = .data[[y_var]], fill = Group)) +
    # Boxplot
    geom_boxplot(alpha = 0.5,
                 outlier.shape = 21, outlier.size = 2) +
    # Significance testing
    geom_signif(
      comparisons = list(levels(data$Group)),  
      map_signif_level = function(p) "*",    
      y_position = y_max * 1.01,
      tip_length = 0,
      textsize = 18,
      size = 1.3
    ) +
    scale_y_continuous(limits = c(NA, y_max * 1.14)) +
    theme_bw() +
    labs(x = "", y = y_label) +
    theme(
      axis.title = element_text(size = 21),
      axis.text = element_text(size = 18),
      legend.position = "none"
    )
}

# Plot and save Succinic acid semialdehyde levels
p_Succinic <- plot_unified(Succinic_acid_semialdehyde, "Succinic acid semialdehyde", "Succinic acid semialdehyde(μg/mg)")
ggsave(p_Succinic, filename = path_target("Succinic acid semialdehyde.png"), width = 6, height = 6)
ggsave(p_Succinic, filename = path_target("Succinic acid semialdehyde.pdf"), width = 6, height = 6)
```


## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r list-files-target}
projthis::proj_dir_info(path_target(), tz = "CET")
```
